---
title: "Appendix 4: example of index standardization of fishery-independent survey data with inlabru"
output:
  pdf_document:
    highlight: tango
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE
)
```


```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
library(INLA)
library(inlabru)
theme_set(theme_minimal())
```

This model will attempt to replicate the model fit in Appendix 2 as an example of how index standardization can also be achieved with inlabru. 
Once again we will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
The density units are kg/km^2^. 
This can be converted to tonnes per km2 by dividing by 1000, which can sometimes scale better.

```{r}
pcod$density_tpkm <- pcod$density/1000
```

We will start by building an identical mesh using INLA.
Here, X and Y are coordinates in UTM zone 9.
These same functions that are used internally in the sdmTMB `make_mesh()` function. 

```{r mesh, echo = TRUE, cache=FALSE}
loc_xy <- as.matrix(pcod[, c("X", "Y"), drop = FALSE])
inla_mesh <- INLA::inla.mesh.create(loc_xy, refine = TRUE, cutoff = 10)

spde <- inla.spde2.pcmatern(
  mesh = inla_mesh,
  prior.range = c(10, 0.05), # P(range < 10) = 0.05
  prior.sigma = c(2, 0.05) # P(sigma > 2) = 0.05
) 
```

```{r plot-mesh, fig.asp=0.9, out.width="4in"}
plot(spde$mesh)
```

There are a couple ways to include the fixed effects with `inlabru`. Unfortunately, using factor variables in the standard equation syntax does not work. In the first case, we can create a design matrix and add year effects individually to the formula.

Now we can set up the design matrix for `inlabru`  

```{r fit-pcod, warning=FALSE, message=FALSE}
m <- lm(density_tpkm ~ -1 + as.factor(year), data = pcod)
model_mat <- model.matrix(m)
colnames(model_mat) <- paste0("x", unique(pcod$year))
dat <- cbind(pcod, model_mat)
```

Next we set up the `components` object for the model. This includes a formula with fixed or random effects (here only fixed) and the design of the spatial random field. The name `spatrf` can be anything, but includes the coordinates, spde model, and how the year effects are shared. 

```{r}
components <- density_tpkm ~ -1 + x2003 + x2004 + x2005 + x2007 +
  x2009 + x2011 + x2013 + x2015 + x2017 +
  # spatrf(main = coordinates, model = spde)
  spatrf(main = coordinates, model = spde, group=year, ngroup=ncol(model_mat),
      control.group = list(model = "iid"))

fit <- bru(
  components = components,
  family = "tweedie",
  # family = "binomial",
  options = bru_options(bru_verbose = TRUE), 
  data = dat
)
bru_coefs <- fit$summary.fixed[, c("mean", "sd")]
```

We can also fit the model with `sdmTMB`. Note that this is a model with only the fixed year effects and a single spatial field, with presence-absence as the response.  

```{r sdmtmb}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)

sdmtmb_fit <- sdmTMB(density_tpkm ~ -1 + as.factor(year),
  data = pcod,
  time = "year", mesh = mesh,
  family = tweedie(link = "log"),
  # family = binomial(link = "logit"),
  # priors = sdmTMBpriors(
  #   matern_st = pc_matern(
  #     range_gt = 10, range_prob = 0.05,
  #     sigma_lt = 2, sigma_prob = 0.05
  #   )
  # ),
  spatiotemporal = "IID"
)
sdmtmb_coefs <- tidy(sdmtmb_fit, effects = "fixed")
```

Comparing the fixed effects between `sdmTMB` and `inlabru` shows the estimates are correlated, but not perfectly so  

```{r}
plot(sdmtmb_coefs$estimate, bru_coefs$mean)
abline(a = 0, b = 1)
```

The other way to set up the fixed coefficients with `inlabru` is to use the `fac` argument. Note the location of the parameters changes from to `summary.random$fac`

```{r}
# components <- present ~ 0 +
#   fac(main = year, model = "factor_full") +
#   spatrf(main = coordinates, model = spde)
# 
# fit2 <- bru(
#   components = components,
#   family = "binomial", data = dat
# )
# bru_coefs2 <- fit2$summary.random$fac[, c("mean", "sd")]
```

These estimates are perfectly correlated with the estimates above,

```{r}
# plot(bru_coefs$mean, bru_coefs2$mean)
```
