---
title: Detailed model description
output:
  bookdown::pdf_document2:
  toc: false
number_sections: false
fig_caption: true
highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
---

\newcommand{\s}{\boldsymbol{s}}

## Notation conventions

This appendix uses the following notation conventions:

* Bold lowercase for vectors

* Bold subscript $\s$ since x, y

* Bold uppercase for matrices

* $\phi$ for all 'dispersion' parameters for consistency with code

* Family titles link to TMB or sdmTMB source code

* Attempt to link math symbols to argument options in model and code symbols


```{r, results='asis'}
symbols <- tibble::tribble(
  ~Symbol, ~Description,
  "$\\phi$",   "A dispersion parameter for a distribution",
  "$\\mu$",   "Mean",
  "$f$",   "Link function"
)
knitr::kable(symbols, caption = "Symbol notation and descriptions", escape = FALSE, format = "latex",  booktabs = TRUE)
```

```{r}
subscripts <- tibble::tribble(
  ~Symbol, ~Description,
  "$\\s$",   "Index for space; a vector of x and y coordinates",
  "$t$",   "Index for time",
  "$g$",   "Group"
)
knitr::kable(subscripts, caption = "Subscript notation", escape = FALSE, format = "latex",  booktabs = TRUE)
```


