\clearpage

# Speed testing related packages {#app:speed}

Here, we describe the methods underlying the speed testing in Fig.\ \ref{fig:timing}.
We generated a mesh that was consistent across simulated data sets for a given mesh resolution (Fig.\ \ref{fig:mesh-vis-timing}).
We did this by setting the `max.edge` argument, which controls the largest allowed triangle edge length.
We tested values of `max.edge` of 0.06, 0.075, 0.1, 0.15, and 0.2.
In Figure 4, we report on the x-axis the number of mesh nodes ("knots" or "vertices") that result from each of these meshes.

We simulated 1000 spatial observations with both x and y coordinates from uniform(0, 1) distributions (Fig.\ \ref{fig:sim-dat-plot-timing}).
Each iteration generated unique coordinates, predictor data, Gaussian random field values, and observation error.
The Gaussian random field was parameterized with a range of 0.5 and marginal standard deviation of 0.2.
The model included an intercept with a value of 0.2 and a normal(0, 1) predictor with an associated coefficient of -0.4.
The observation error was Gaussian with a standard deviation of 0.3.

We conducted 50 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values (Fig.\ \ref{fig:timing}).
The models were fit on a 2022 14 inch M1 MacBook Pro with an Apple M1 Pro 10-core CPU and 32 GB of RAM in R 4.2.2 and the default Matrix library. The following package versions were used:

```{r pkg-versions, echo=TRUE}
packageVersion("INLA")
packageVersion("inlabru")
packageVersion("mgcv")
packageVersion("spaMM")
packageVersion("sdmTMB")
```

\subsection[Illustration of generating the INLA mesh]{Illustration of generating the \pkg{INLA} mesh}

```{r pkgs, warning=FALSE, message=FALSE, cache=FALSE, echo=FALSE}
library("INLA")
library("inlabru")
library("ggplot2")
library("sdmTMB")
library("mgcv")
library("spaMM")
```

We will illustrate with a `max.edge` of 0.06:

```{r mesh-timing, echo=TRUE}
max_edge <- 0.06
loc_bnd <- matrix(c(0, 0, 1, 0, 1, 1, 0, 1), 4, 2, byrow = TRUE)
segm_bnd <- INLA::inla.mesh.segment(loc_bnd)
mesh <- INLA::inla.mesh.2d(
  boundary = segm_bnd,
  max.edge = c(max_edge, 0.2),
  offset = c(0.1, 0.05)
)
```

This mesh has `r mesh$n` (`mesh$n`) vertices.

```{r mesh-vis-timing, fig.width=7, fig.asp=1.1, echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='ht', fig.align='center'}
max_edge_vec <- rev(c(0.06, 0.075, 0.1, 0.15, 0.2))
g <- list()
for (i in seq_along(max_edge_vec)) {
  .mesh <- INLA::inla.mesh.2d(
    boundary = segm_bnd,
    max.edge = c(max_edge_vec[i], 0.2),
    offset = c(0.1, 0.05)
  )
  g[[i]] <- ggplot() +
    gg(.mesh) +
    theme_light() +
    coord_equal(expand = FALSE) +
    ggtitle(paste("Mesh vertices =", .mesh$n))
}
cowplot::plot_grid(plotlist = g, ncol = 2)
```

\clearpage

## Illustration of simulating data

```{r sim-timing, echo=TRUE}
set.seed(123)
n_obs <- 1000
predictor_dat <- data.frame(
  X = runif(n_obs), Y = runif(n_obs),
  a1 = rnorm(n_obs)
)
mesh_sdmTMB <- make_mesh(predictor_dat, xy_cols = c("X", "Y"), mesh = mesh)

sim_dat <- sdmTMB_simulate(
  formula = ~ 1 + a1,
  data = predictor_dat,
  mesh = mesh_sdmTMB,
  family = gaussian(),
  range = 0.5,
  phi = 0.3,
  sigma_O = 0.2,
  B = c(0.2, -0.4) # B0 = intercept, B1 = a1 slope
)
```

<!-- Look at the simulated data: -->

```{r sim-dat-plot-timing, out.width="5in", echo=FALSE, fig.cap="Example simulated dataset with a spatial random field.", fig.pos='ht', fig.align='center'}
ggplot(sim_dat, aes(X, Y, colour = observed, size = abs(observed))) +
  geom_point() +
  scale_color_gradient2() +
  coord_fixed() +
  theme_light()
```

\subsection[Example sdmTMB model fit]{Example \pkg{sdmTMB} model fit}

```{r sdmTMBfit-timing, echo=TRUE}
fit_sdmTMB <- sdmTMB(
  observed ~ a1,
  data = sim_dat, 
  mesh = mesh_sdmTMB, 
  family = gaussian(),
  priors = sdmTMBpriors(
    matern_s = pc_matern(range_gt = 0.05, sigma_lt = 2)
  )
)
```

\subsection[Example spaMM model fit]{Example \pkg{spaMM} model fit}

```{r spaMMfit-timing, echo=TRUE, cache=TRUE}
spde <- INLA::inla.spde2.pcmatern(
  mesh = mesh,
  prior.range = c(0.05, 0.05),
  prior.sigma = c(2, 0.05)
)

fit_spaMM <- fitme(
  observed ~ a1 + IMRF(1 | X + Y, model = spde),
  family = gaussian(), 
  data = sim_dat
)
```

\subsection[Example inlabru model fit]{Example \pkg{inlabru} model fit}

```{r inlabrufit-timing, echo=TRUE, cache=TRUE}
# convert to sp SpatialPointsDataFrame first:
dat_sp <- sp::SpatialPointsDataFrame(
  cbind(sim_dat$X, sim_dat$Y),
  proj4string = sp::CRS(
    "+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000
+ +y_0=0 +datum=NAD83 +units=km +no_defs"
  ), data = sim_dat
)
components <- observed ~ -1 + Intercept(1) + a1 +
  spatrf(main = coordinates, model = spde)
like <- like(observed ~ Intercept + a1 + spatrf,
  family = "gaussian", data = dat_sp
)
fit_bru <- bru(
  like,
  components = components,
  options = bru_options(
    control.inla = list(int.strategy = "eb", strategy = "gaussian"),
    bru_max_iter = 1, num.threads = "1:1"
  )
)
```

\subsection[Example mgcv model fit]{Example \pkg{mgcv} model fit}

```{r mgcv-spde-funcs-timing, echo=FALSE, eval=TRUE}
# -------------------------------------------------------------------------
# Code in this chunk is from:
#
# Miller, D.L., Glennie, R. & Seaton, A.E. Understanding the Stochastic Partial
# Differential Equation Approach to Smoothing. JABES 25, 1–16 (2020).
# https://doi.org/10.1007/s13253-019-00377-z
#
# Re-used here under a Creative Commons Attribution 4.0 International License
# http://creativecommons.org/licenses/by/4.0/
smooth.construct.spde.smooth.spec <- function(object, data, knots) {
  dim <- length(object$term)
  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
  if (dim == 1) {
    x <- data[[object$term]]
  } else {
    x <- matrix(0, nr = length(data[[1]]), nc = 2)
    x[, 1] <- data[[object$term[1]]]
    x[, 2] <- data[[object$term[2]]]
  }
  if (is.null(object$xt)) {
    if (dim == 1) {
      t <- seq(min(x), max(x), len = object$bs.dim)
      mesh <- INLA::inla.mesh.1d(loc = t, degree = 2, boundary = "free")
    } else {
      stop("For 2D, mesh must be supplied as argument xt$mesh in s(...,xt = )")
    }
  } else {
    if (class(object$xt$mesh) != "inla.mesh") stop("xt must be NULL or an inla.mesh object")
    mesh <- object$xt$mesh
  }
  object$X <- as.matrix(INLA::inla.spde.make.A(mesh, x))
  inlamats <- INLA::inla.mesh.fem(mesh)
  object$S <- list()
  object$S[[1]] <- as.matrix(inlamats$c1)
  object$S[[2]] <- 2 * as.matrix(inlamats$g1)
  object$S[[3]] <- as.matrix(inlamats$g2)
  object$L <- matrix(c(2, 2, 2, 4, 2, 0), ncol = 2)
  object$rank <- rep(object$bs.dim, 3)
  object$null.space.dim <- 0
  object$mesh <- mesh
  object$df <- ncol(object$X)
  class(object) <- "spde.smooth"
  return(object)
}
Predict.matrix.spde.smooth <- function(object, data) {
  dim <- length(object$term)
  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
  if (dim == 1) {
    x <- data[[object$term]]
  } else {
    x <- matrix(0, nr = length(data[[1]]), nc = 2)
    x[, 1] <- data[[object$term[1]]]
    x[, 2] <- data[[object$term[2]]]
  }
  Xp <- INLA::inla.spde.make.A(object$mesh, x)
  return(as.matrix(Xp))
}
# End of code from
# Miller, D.L., Glennie, R. & Seaton, A.E. Understanding the Stochastic Partial
# Differential Equation Approach to Smoothing. JABES 25, 1–16 (2020).
# https://doi.org/10.1007/s13253-019-00377-z
#
# Re-used here under a Creative Commons Attribution 4.0 International License
# http://creativecommons.org/licenses/by/4.0/
# -------------------------------------------------------------------------
```

First define `smooth.construct.spde.smooth.spec()` and `Predict.matrix.spde.smooth()`
from the supplement of @miller2019, then:

```{r mgcvfit-timing, echo=TRUE, eval=TRUE}
fit_bam <- bam(
  observed ~ a1 + s(X, Y,
    bs = "spde", k = mesh$n,
    xt = list(mesh = mesh)
  ),
  data = sim_dat,
  family = gaussian(),
  method = "fREML",
  control = gam.control(scalePenalty = FALSE), 
  discrete = TRUE
)
```

\clearpage
